The Talking Heads Experiment

### 2015-04-27 chap2/chap2.tex
fixed index entries and extra papers, added proofreaders and typesetters

Let me make a distinction between the physical aspects of 
a cognitive agent and the mental aspects. The physical aspects include the 
agent's body, the sensors and articulators, 
the physical location, the objects in this location, and the other
agents physically present in the environment. The 
mental aspects include the agent's repertoire of behaviours, 
the brain structures and processes
performing categorisations of reality, his
memory, lexicon, grammar, and so on. 
In the case of humans, these two aspects are intimately 
connected and indivisible. We cannot teleport our mental 
faculties into another body, or into another copy of 
our body, even though there have been speculations that 
we could in the future record human 
brain states,\footnote{
Such visions of the future have been put forward
by \cite{Moravec:1995}. Neither current artificial intelligence technology nor
the state of the art in brain state recording are anywhere 
near to realising these visions.}
I believe that this will still not enable teleportation
because in human brains there is no distinction between 
hardware and software. The architecture of a human brain, 
the physical connections between cells, and the 
biochemical processes in each cell determine the brain's behaviour.
There is no separation between a {\itshape brain program} and
an interpreter that reads brain programs and 
executes them. The brain is a special-purpose 
hardware device which is unique to each individual. 
To copy such a device we would have to rebuild it physically,
atom by atom, and integrate it in an exact copy of the 
same body.

@@

In constructing the robot bodies, we have used as much as
possible off-the-shelf standard 
components so that we could focus almost completely on issues
directly relevant to language and meaning.
Each robot's low-level vision system
(integrated in the camera) is already very sophisticated.\footnote{
The camera is a Sony EVI-D31. The main computer is a Power 
Macintosh from Apple, Inc. The agent servers run under the 
Linux operating system.}
It can focus automatically to get a sharper image
and autonomously track a moving object. The speech signal is 
produced with a standard text-to-speech system 
so that we did not have to 
worry about building complex audio modules ourselves.
The computer hardware 
is powerful, but not specialised nor in the supercomputer 
range. All programs have been written in the standard symbolic 
programming language of artificial intelligence 
research, namely \textsc{lisp}.\footnote{
The construction of the programs underlying the 
Talking Heads experiment has required advanced
artificial intelligence programming techniques, 
such as discussed by \cite{Norvig:1996}. A general
toolkit for the systematic execution of 
simulation and physical experiments, called \textsc{babel}, has been
designed and implemented by Angus McIntyre. The 
toolkit allows the definition and modular composition
of cognitive architectures, the design of 
experiments, and the monitoring and displaying of 
results, see \cite{McIntyre:1998}.} What makes the Talking Heads experiment
special is not the hardware or software tools
but what we have done with it.

@@

Given this rich computational and robotic infrastructure, many \is{Guessing Game}
specific experiments are possible. 
Each experiment could explore
a particular interaction between the robots, the environment, 
and human observers. In this book, I explore only 
one type of interaction, which I call the
{\itshape Guessing Game}.\footnote{
A very similar game has been called the original 
language game by \cite{Brown:1973} in the context of
research on child language acquisition. See also 
the thoughtful analysis in \cite{Halliday:1975}. 
Research on child language has inspired the 
agent architectures and behaviours but they should not
be seen as a realistic model of child language acquisition.}
The Talking Heads play this game 
either among themselves or with a human experimenter.

@@

The speaker first looks at one area of the white board and 
directs the attention of the hearer to the same area.\footnote{
I will explain later how exactly agents
decide on a particular scene and how they are able to 
draw each other's attention to specific areas of the 
white board.}
The objects located in this area constitute the {\itshape context}. 
The speaker then chooses one object from the context, which 
I will call the {\itshape topic}, and gives a verbal hint
to the hearer. The {\itshape verbal hint} is an expression
that identifies the topic 
with respect to the other objects in the context. For example, 
if the context contains a red square, a blue triangle, 
and a green circle, then the speaker may say something like 
`the red one' to identify the red square as the topic. 
If the context contains also a red triangle, he has to be more 
precise and say something like `the red square' to delineate
the topic from the red triangle as well as
from the blue square. 
Of course, the Talking Heads do not say `the red square'
but use their own language and concepts which are not 
necessarily the same as those used in English. For example, 
they may say `malewina' to mean [\textsc{upper extreme}-\textsc{left low}-\textsc{redness}].

@@

The game is different from a closed world game like 
chess, because the environment is open. The human 
experimenter may introduce new objects at any
time, any one agent can extend
the language (for example invent a new word), possibly
requiring the other agents to adopt it as well, or the
human experimenter may inject new language forms in the 
dialogues. Agents try to maximise their communicative 
success by cooperating to the fullest and update or change
their internal structures and processes \enlargethispage{1\baselineskip}
to improve their chances in the game.\footnote{
The Guessing Game is a cooperative game, because
both the agents win or loose at the same time and have 
the highest gain if they develop co-ordinated behaviour. 
The agent who manages to have the most success in the game is 
the global winner. Because the commentator requires to know 
from the speaker which topic he wants to communicate 
before he is allowed to speak, no cheating 
is possible. Game theory, originally founded by John 
von Neumann and Oskar Morgenstern, can be applied to study the 
language game mathematically. \is{game theory} We are dealing with an 
evolutionary game in which the players optimise their 
internal states to become better in the game, see
\cite{Maynard:1982}.}

In the Talking Heads experiment, it is assumed that 
agents want to cooperate and that they use communication
as part of their cooperation. 
The evolution of cooperative games has been studied 
extensively by artificial life researchers often 
in the context of Robert Axelrod's prisoners dilemma game,\is{prisoners dilemma game} 
see for example: \cite{Ikegami:1994}.  
For a general introduction how communication can 
evolve in the context of cooperation, see: \cite{Hauser:1996}. 
Computer simulations showing the evolution of 
cooperation and communication have been reported in the
artificial life literature. See for example: \cite{MacLennan:1991}. 
Most of these computer simulations are 
closer to animal signaling systems than to human lexicons
both in size and in terms of the complexity of meaning.

@@

The environment of the Talking Heads is 
not fixed. The human experimenter 
may change the position of objects, add new kinds of 
objects, or eliminate
others. Consequently a strategy of naming individual objects
will not work. It would lead to a proliferation of 
proper names and it would require the Talking Heads to
recognise objects, which is very difficult to do.\footnote{
For a thorough exposition of the difficulties of 
object recognition, see \cite{Ullman:1996}. 
Object constancy comes fairly late in the 
acquisition of a child's ontology, as Piaget's conservation
experiments have shown. Language probably plays an important
role in forming the notion of an object.}
Indeed, humans don't exclusively use proper names in 
natural language conversations 
either. We say `could you give me the red small square' as 
opposed to `could you give me O\_143'. Natural language
words like `red' or `small' name perceptually grounded categories
and syntactic structures indicate how they should
be combined and used to find the topic. The relation between 
a language expression and its referent is therefore 
always indirect. This is summarised
in the semiotic square (\figref{triangle}), which I will use throughout this book to 
help understand and analyse the nature of language communication. \is{semiotic square}
The semiotic square relates the four entities
involved in a verbal interaction: 
\begin{itemize}
\item An {\itshape utterance}, such as `small red square', which is 
transmitted as a physical 
signal from one agent to another one through the 
external environment. It is written between double quotes. 
\item A {\itshape meaning}, which consists of categories like [\textsc{red}], [\textsc{small}], 
or combinations of categories, like \{[\textsc{red}] [\textsc{small}]\}. Labels of 
categories are written in capital letters between 
square brackets. 
\clearpage
\item An {\itshape image segment}, denoted as \emph{S143}, which is 
a segment of the image perceived through the camera.
\item A {\itshape referent}, like object \emph{O143}, which is 
an entity in the real world. 
\end{itemize}

@@

The systematic relation between meaning and referent is 
usually studied under the heading of semantics, and the 
systematic relation between meaning and utterance as
grammar (including syntax, morphology and lexicon).\footnote{
For a general introduction to the contemporary linguistic
viewpoint on the processes involved, see: \cite{Vanvalin:1997}. 
In a logical approach to language, as 
exemplified by Montague grammar (\citealt{Montague:1974}),  
meanings are represented using a logical formalism, i.e.
a variant of intensional logic. \is{Montague grammar}
Natural language expressions are systematically 
related to expressions in this logic, and a
formal semantics system defines how expressions in the logic
are mapped onto their denotations. Because this is 
a formal framework, the denotations consist of formal 
models. To make the Talking Heads experiment work, 
we needed to develop a grounded semantics system, which details
how an agent may go from physical reality to meaning
using a perceptual apparatus, and from meaning 
to physical reality. The logical structure of 
the meanings we will investigate are very simple 
(unary predicates and conjunctions). But once we know how to 
set up a grounded semantics for simple meaning
structures we can scale it up to the more complex meaning
structures typically studied in logic.}

Many tricky philosophical issues are raised in 
the unavoidable distinction between the sensed image
of an object (which is local to the agent)
and the object itself (which is external to the 
agent). Some philosophers even doubt that objects 
have an existence outside our perception of them! 
We need to make the distinction because agents always
have different internal images even if they look at 
the same object seen from our viewpoint as an external
observer. However, for simplifying the 
explanations, I will sometimes assume that perceived
image and external object are the same, so that the 
semiotic square becomes a semiotic triangle.\footnote{
The notion of a semiotic triangle was first
introduced in a classic of the semiotic literature, 
see \cite{Ogden:1935}.}

@@

I deliberately left an important aspect of language out. 
In the case of physical agents, the form cannot be transmitted
directly but needs to be articulated in speech sounds, 
written signs, or gestures, to create a true utterance. This additional 
complexity will not be discussed further in this book -- even though 
it is a fascinating topic in its own right.\footnote{
See for a state of the art review: \cite{Clark:1990}. 
We have already been conducting quite extensive 
research in our group on the origin of sound systems 
using similar principles as the ones discussed in this 
book for the origins of word semantics. 
See: \cite{DeBoer:1997}. The complex 
adaptive systems approach underlying this phonetics
work was already foreshadowed in work of phoneticians
like \cite{Liljencrants:1972}.}
To make a verbal interaction nevertheless complete, agents
are given a repertoire of consonants and vowels with 
which they can make random syllables and syllable combinations, 
like `wabido', `bimaku', etc. The articulation and 
recognition of these syllables is assumed to be acquired
already and transmission is engineered to be error-free. 
This way, our attention can be focused on how
ontologies, lexicons, and grammars may emerge.

@@

Segmentation can happen according to several criteria. \is{image segmentation}
For example, patches with similar colour can be grouped 
into a single patch, edges can be identified 
and then linked with each other to form line segments
circumscribing the contours of 
an object, or consecutive images can be compared to extract the parts
that changed and hence moved as a single object. 
There is now a solid body of techniques from decades
of machine vision research to efficiently segment scenes according to 
these and other methods. The Talking Heads use several methods
in parallel and combine their output to get a clearly segmented
picture.\footnote{
For general introductions to these areas, 
see: \cite{Ballard:1982} and \cite{Fischler:1987}.}

@@

There are obviously other ways to 
move from the continuous domain of sensory channels to 
the discrete domain of categories.\footnote{
See: \cite{Taylor:1989}.}
For example, it is not really necessary 
to have a binary division, we could just as well split 
each region into three or more subregions. Or we could introduce
focal (prototypical) values and associate
a category with each of them. In the latter case, 
the categorisation process consists 
in identifying the prototype that is closest to 
an object's value. For the current experiment, I will stick however
to a binary categorisation strategy because that is
the simplest to understand and formally investigate.

@@

In the early origins of language,
there must have been an initial phase in which no complex syntax 
was in place yet. Utterances then must have consisted of single words
or multiple words without further syntax. \is{proto-language}
Such syntax-less languages have been called {\itshape proto-languages}.\footnote{
See \cite{Bickerton:1990} as well as \cite{Thomason:1988}. Children similarly 
go through a single word phase (even though a single 
word for them might be multiple words for us), and 
then slowly start to bootstrap their grammar. 
See: \cite{Tomasello:1991} and \cite{Bates:1991}. 
They are still observed in the very first phases of 
child language acquisition or in pidgins that spontaneously 
form when two communities with widely diverging languages
need to interact. Out of proto-languages, languages 
with a fully-fledged syntax must have emerged at some point. 
How this occurred is still a heavily debated mystery. In this volume 
I only treat the origins of proto-languages.}

Children acquire their first words around the first year
of life.\footnote{
Representative work in the study of child language
learning focuses mostly on the acquisition of specific meanings.
See: \cite{Gleitman:1994}, 
\cite{Clark:1993}, \cite{Bowerman:1996}.}
Most people believe that they do this 
as a result of hearing a particular word repeated several times in a
certain context and gradually abstracting an
association between a word and a meaning. But 
how is it that they know which meaning to associate with a 
particular word? How is it that word acquisition goes
so rapid? We will follow a different 
approach, which leaves an open question whether
this applies to human word acquisition as well. The
approach will be selectionist. The agents {\itshape construct} 
hypotheses, either on the basis of one specific case 
where they guess through a non-verbal strategy what 
the meaning of an unknown word might be, 
or they have simply invented a new word because
they do not have one yet. The hypotheses are 
then tried out in subsequent games and either receive confirmation
or are falsified. As a side effect of this local behaviour, 
a global self-organising dynamic process arises leading gradually to coherence.

@@

Through such disambiguating situations, meanings of words
get clarified and the lexicons of the agents
become more similar. Note that the 
dominating meaning of `lu' can still go in 
two directions. For example, {\bfshape a2}
could have used `lu' with {\bfshape a1} 
in a situation where its only possible meaning 
is [\textsc{hpos} 0.0–0.5] (left). Then {\bfshape a1}, now playing the role
of hearer, would have stored the association between
`lu' and [\textsc{hpos} 0.0–0.5]. If this happened often 
enough [\textsc{hpos} 0.0–0.5] (left) might
have become the dominant meaning of `lu', instead of 
[\textsc{vpos} 0.0–0.5] (lower). This shows clearly that the 
evolution of language will never be predictable. At a critical 
bifurcation, small preferential differences between 
the agents or the chance occurrence of certain situations
may tilt the competition in one direction or the other.\footnote{
This high sensitivity to initial conditions is 
one of the characteristics of a chaotic dynamic
system, see \cite{Lorenz:1993}.
Indeed, as I will explore in more detail later, 
evolving languages have all the characteristics
of complex adaptive systems, including the 
potential for punctuated equilibria.}
There is no right
or wrong solution in the language game and no one has
any more rights than anyone else.

@@

Another major difference with classical theories of 
meaning and reference is 
that the repertoire of meanings and form-meaning pairs is open-ended
and subject to change at any time. Logicians would say that the 
truth conditions of the forms are non-monotonic. \is{non-monotonicity}
Non-monotonicity is unavoidable in the case of grounded 
situated agents with limited rationality.\footnote{
Examples of these processes are described in: 
\cite{Bybee:1994}, \cite{Traugott:1991}.
For attempts to explain these ``grammaticalisation'' 
processes in terms of general cognitive operations, see: 
\cite{Heine:1997}.}
Agents should always be allowed to 
introduce new forms or recruit existing forms to express new
meanings, simply because the set of meanings must expand
to cope with novel contexts and new communicative
situations. After enough interactions among the 
members of the same group in a relatively stable 
environment, we expect there to be
a stable set of conventionalised form-meaning 
relations, but it cannot be
expected that everything anyone ever wants to say is
already conventionalised, and so there will always
be turbulence at the fringes. The idea that the lexicon
and the syntax of a language are static
entities is completely false. Both are in constant flux.
Non-monotonicity is consequently one of the big 
topics in logic-based approaches to artificial intelligence.

@@

The spontaneous generation of repertoires of complex 
semantic strategies already explains some characteristics 
that are reflected in full-fledged languages: 
There is hierarchical structure, because one strategy 
may call upon another strategy to achieve a subgoal, 
and a strategy may potentially call upon itself thus introducing 
recursivity. There is also a functional specialisation
of categories. They are now sometimes used for filtering
the members of a set, sometimes for ordering them, sometimes
for modifying another category before it is used for 
filtering, and so on. Hierarchical structuring, recursivity
and functional specialisation therefore do not have to 
originate in language. The fact that we see them in 
language structure is a reflection of the generic 
hierarchical and functional nature 
of semantic strategies.\footnote{
The fact that languages change is, of course, 
well-known in linguistics even though it is less a focal 
research topic than used to be in the 19th century. 
See: \cite{McMahon:1994} for a brief introduction into the 
subject. Language change has many of the characteristics
of biological change but it takes place in cultural 
as opposed to biological time. See: \cite{Labov:1994}.}

Second, my hypothesis is that the need for communicating 
more and more complex semantic strategies and the 
conceptual structures they use, has pushed language 
users to start using more and more complex form characteristics, 
starting with word order but also intonation, stress 
patterns, function words, morphological variations of words, 
etc. Here another aspect comes into 
play: the ability to recognise forms and structures, 
which we also need to recognise structured objects in 
scenes for example, and the ability to assemble a 
structure satisfying a set of constraints. The collective
dynamics which are responsible for the propagation and the spontaneous 
self-organisation of coherence in lexicons should also apply  
here. Grammars can be seen as ecologies, where
form-meaning pairs compete in the population. New 
syntactic and semantic categories, new constructions and 
new uses of grammatical conventions are continuously created,
and existing ones may destabilise and become
in disuse.\footnote{
The semantic strategies are similar to the 
cognitive processes that have been studied extensively 
in cognitive grammar (\citealt{Langacker:1987}). Formal versions
of some of these processes have been formulated as 
part of Montague grammar (\citealt{Montague:1974}). In 
computational linguistics research, various 
attempts have been made to formalise conceptualisation
and meaning application strategies (\citealt{Gazdar:1989}).}

@@

\enlargethispage{1\baselineskip}
According to this theory, natural language structure is
not the consequence of an 
autonomous innate language device which evolved 
by a genetic mutation or series of mutations. Instead 
cognitive mechanisms and structures already in place have been 
marshalled to get a language system off the ground, even though 
once language became essential for the rapid incorporation
of new individuals and the general organisation of 
activities in human populations, it must have started
to recruit vast amounts of brain capacity and stimulated 
other cognitive faculties, such as categorisation,
episodic memory or problem solving, to become in turn 
much more complex and versatile.\footnote{
There has been an ongoing nature versus nurture debate 
with respect to the origins and acquisition of language. 
The different issues in this debate are already well
illustrated in \cite{Piattelli:1980}.}

>>>>

Let me make a distinction between the physical aspects of 
a cognitive agent and the mental aspects. The physical aspects include the 
agent's body, the sensors and articulators, 
the physical location, the objects in this location, and the other
agents physically present in the environment. The 
mental aspects include the agent's repertoire of behaviours, 
the brain structures and processes
performing categorisations of reality, his
memory, lexicon, grammar, and so on. 
In the case of humans, these two aspects are intimately 
connected and indivisible. We cannot teleport our mental 
faculties into another body, or into another copy of 
our body, even though there have been speculations that 
we could in the future record human 
brain states,\footnote{Such visions of the future have been put forward
by \cite{Moravec:1995}. Neither current artificial intelligence technology nor
the state of the art in brain state recording are anywhere 
near to realising these visions.}
I believe that this will still not enable teleportation
because in human brains there is no distinction between 
hardware and software. The architecture of a human brain, 
the physical connections between cells, and the 
biochemical processes in each cell determine the brain's behaviour.
There is no separation between a {\itshape brain program} and
an interpreter that reads brain programs and 
executes them. The brain is a special-purpose 
hardware device which is unique to each individual. 
To copy such a device we would have to rebuild it physically,
atom by atom, and integrate it in an exact copy of the 
same body.

@@

In constructing the robot bodies, we have used as much as
possible off-the-shelf standard 
components so that we could focus almost completely on issues
directly relevant to language and meaning.
Each robot's low-level vision system
(integrated in the camera) is already very sophisticated.\footnote{The camera is a Sony EVI-D31. The main computer is a Power 
Macintosh from Apple, Inc. The agent servers run under the 
Linux operating system.}
It can focus automatically to get a sharper image
and autonomously track a moving object. The speech signal is 
produced with a standard text-to-speech system 
so that we did not have to 
worry about building complex audio modules ourselves.
The computer hardware 
is powerful, but not specialised nor in the supercomputer 
range. All programs have been written in the standard symbolic 
programming language of artificial intelligence 
research, namely \textsc{lisp}.\footnote{The construction of the programs underlying the 
Talking Heads experiment has required advanced
artificial intelligence programming techniques, 
such as discussed by \cite{Norvig:1996}. A general
toolkit for the systematic execution of 
simulation and physical experiments, called \textsc{babel}, has been
designed and implemented by Angus McIntyre. The 
toolkit allows the definition and modular composition
of cognitive architectures, the design of 
experiments, and the monitoring and displaying of 
results, see \cite{McIntyre:1998}.} What makes the Talking Heads experiment
special is not the hardware or software tools
but what we have done with it.

@@

Given this rich computational and robotic infrastructure, many \is{Guessing Game}
specific experiments are possible. 
Each experiment could explore
a particular interaction between the robots, the environment, 
and human observers. In this book, I explore only 
one type of interaction, which I call the
{\itshape Guessing Game}.\footnote{A very similar game has been called the original 
language game by \cite{Brown:1973} in the context of
research on child language acquisition. See also 
the thoughtful analysis in \cite{Halliday:1975}. 
Research on child language has inspired the 
agent architectures and behaviours but they should not
be seen as a realistic model of child language acquisition.}
The Talking Heads play this game 
either among themselves or with a human experimenter.

@@

The speaker first looks at one area of the white board and 
directs the attention of the hearer to the same area.\footnote{I will explain later how exactly agents
decide on a particular scene and how they are able to 
draw each other's attention to specific areas of the 
white board.}
The objects located in this area constitute the {\itshape context}. 
The speaker then chooses one object from the context, which 
I will call the {\itshape topic}, and gives a verbal hint
to the hearer. The {\itshape verbal hint} is an expression
that identifies the topic 
with respect to the other objects in the context. For example, 
if the context contains a red square, a blue triangle, 
and a green circle, then the speaker may say something like 
`the red one' to identify the red square as the topic. 
If the context contains also a red triangle, he has to be more 
precise and say something like `the red square' to delineate
the topic from the red triangle as well as
from the blue square. 
Of course, the Talking Heads do not say `the red square'
but use their own language and concepts which are not 
necessarily the same as those used in English. For example, 
they may say `malewina' to mean [\textsc{upper extreme}-\textsc{left low}-\textsc{redness}].

@@

The game is different from a closed world game like 
chess, because the environment is open. The human 
experimenter may introduce new objects at any
time, any one agent can extend
the language (for example invent a new word), possibly
requiring the other agents to adopt it as well, or the
human experimenter may inject new language forms in the 
dialogues. Agents try to maximise their communicative 
success by cooperating to the fullest and update or change
their internal structures and processes \enlargethispage{1\baselineskip}
to improve their chances in the game.\footnote{The Guessing Game is a cooperative game, because
both the agents win or loose at the same time and have 
the highest gain if they develop co-ordinated behaviour. 
The agent who manages to have the most success in the game is 
the global winner. Because the commentator requires to know 
from the speaker which topic he wants to communicate 
before he is allowed to speak, no cheating 
is possible. Game theory, originally founded by John 
von Neumann and Oskar Morgenstern, can be applied to study the 
language game mathematically. \is{game theory} We are dealing with an 
evolutionary game in which the players optimise their 
internal states to become better in the game, see
\cite{Maynard:1982}.}

In the Talking Heads experiment, it is assumed that 
agents want to cooperate and that they use communication
as part of their cooperation. 
The evolution of cooperative games has been studied 
extensively by artificial life researchers often 
in the context of Robert Axelrod's prisoners dilemma game,\is{prisoners dilemma game} 
see for example \cite{Ikegami:1994}.  
For a general introduction how communication can 
evolve in the context of cooperation, see \cite{Hauser:1996}. 
Computer simulations showing the evolution of 
cooperation and communication have been reported in the
artificial life literature. See for example \cite{MacLennan:1991}. 
Most of these computer simulations are 
closer to animal signaling systems than to human lexicons
both in size and in terms of the complexity of meaning.

@@

The environment of the Talking Heads is 
not fixed. The human experimenter 
may change the position of objects, add new kinds of 
objects, or eliminate
others. Consequently a strategy of naming individual objects
will not work. It would lead to a proliferation of 
proper names and it would require the Talking Heads to
recognise objects, which is very difficult to do.\footnote{For a thorough exposition of the difficulties of 
object recognition, see \cite{Ullman:1996}. 
Object constancy comes fairly late in the 
acquisition of a child's ontology, as Piaget's conservation
experiments have shown. Language probably plays an important
role in forming the notion of an object.}
Indeed, humans don't exclusively use proper names in 
natural language conversations 
either. We say `could you give me the red small square' as 
opposed to `could you give me O\_143'. Natural language
words like `red' or `small' name perceptually grounded categories
and syntactic structures indicate how they should
be combined and used to find the topic. The relation between 
a language expression and its referent is therefore 
always indirect. This is summarised
in the semiotic square (\figref{triangle}), which I will use throughout this book to 
help understand and analyse the nature of language communication. \is{semiotic square}
The semiotic square relates the four entities
involved in a verbal interaction: 
\begin{itemize}
\item An {\itshape utterance}, such as `small red square', which is 
transmitted as a physical 
signal from one agent to another one through the 
external environment. It is written between double quotes. 
\item A {\itshape meaning}, which consists of categories like [\textsc{red}], [\textsc{small}], 
or combinations of categories, like \{[\textsc{red}] [\textsc{small}]\}. Labels of 
categories are written in capital letters between 
square brackets. 
\clearpage
\item An {\itshape image segment}, denoted as \emph{S143}, which is 
a segment of the image perceived through the camera.
\item A {\itshape referent}, like object \emph{O143}, which is 
an entity in the real world. 
\end{itemize}

@@

The systematic relation between meaning and referent is 
usually studied under the heading of semantics, and the 
systematic relation between meaning and utterance as
grammar (including syntax, morphology and lexicon).\footnote{For a general introduction to the contemporary linguistic
viewpoint on the processes involved, see  \cite{Vanvalin:1997}. 
In a logical approach to language, as 
exemplified by Montague grammar (\citealt{Montague:1974}),  
meanings are represented using a logical formalism, i.e.
a variant of intensional logic. \is{Montague grammar}
Natural language expressions are systematically 
related to expressions in this logic, and a
formal semantics system defines how expressions in the logic
are mapped onto their denotations. Because this is 
a formal framework, the denotations consist of formal 
models. To make the Talking Heads experiment work, 
we needed to develop a grounded semantics system, which details
how an agent may go from physical reality to meaning
using a perceptual apparatus, and from meaning 
to physical reality. The logical structure of 
the meanings we will investigate are very simple 
(unary predicates and conjunctions). But once we know how to 
set up a grounded semantics for simple meaning
structures we can scale it up to the more complex meaning
structures typically studied in logic.}

Many tricky philosophical issues are raised in 
the unavoidable distinction between the sensed image
of an object (which is local to the agent)
and the object itself (which is external to the 
agent). Some philosophers even doubt that objects 
have an existence outside our perception of them! 
We need to make the distinction because agents always
have different internal images even if they look at 
the same object seen from our viewpoint as an external
observer. However, for simplifying the 
explanations, I will sometimes assume that perceived
image and external object are the same, so that the 
semiotic square becomes a semiotic triangle.\footnote{The notion of a semiotic triangle was first
introduced in a classic of the semiotic literature, 
see \cite{Ogden:1935}.}

@@

I deliberately left an important aspect of language out. 
In the case of physical agents, the form cannot be transmitted
directly but needs to be articulated in speech sounds, 
written signs, or gestures, to create a true utterance. This additional 
complexity will not be discussed further in this book -- even though 
it is a fascinating topic in its own right.\footnote{See for a state of the art review: \cite{Clark:1990}. 
We have already been conducting quite extensive 
research in our group on the origin of sound systems 
using similar principles as the ones discussed in this 
book for the origins of word semantics. 
See  \cite{DeBoer:1997}. The complex 
adaptive systems approach underlying this phonetics
work was already foreshadowed in work of phoneticians
like \cite{Liljencrants:1972}.}
To make a verbal interaction nevertheless complete, agents
are given a repertoire of consonants and vowels with 
which they can make random syllables and syllable combinations, 
like `wabido', `bimaku', etc. The articulation and 
recognition of these syllables is assumed to be acquired
already and transmission is engineered to be error-free. 
This way, our attention can be focused on how
ontologies, lexicons, and grammars may emerge.

@@

Segmentation can happen according to several criteria. \is{image segmentation}
For example, patches with similar colour can be grouped 
into a single patch, edges can be identified 
and then linked with each other to form line segments
circumscribing the contours of 
an object, or consecutive images can be compared to extract the parts
that changed and hence moved as a single object. 
There is now a solid body of techniques from decades
of machine vision research to efficiently segment scenes according to 
these and other methods. The Talking Heads use several methods
in parallel and combine their output to get a clearly segmented
picture.\footnote{For general introductions to these areas, 
see  \cite{Ballard:1982} and \cite{Fischler:1987}.}

@@

There are obviously other ways to 
move from the continuous domain of sensory channels to 
the discrete domain of categories.\footnote{See  \cite{Taylor:1989}.}
For example, it is not really necessary 
to have a binary division, we could just as well split 
each region into three or more subregions. Or we could introduce
focal (prototypical) values and associate
a category with each of them. In the latter case, 
the categorisation process consists 
in identifying the prototype that is closest to 
an object's value. For the current experiment, I will stick however
to a binary categorisation strategy because that is
the simplest to understand and formally investigate.

@@

In the early origins of language,
there must have been an initial phase in which no complex syntax 
was in place yet. Utterances then must have consisted of single words
or multiple words without further syntax. \is{proto-language}
Such syntax-less languages have been called {\itshape proto-languages}.\footnote{See \cite{Bickerton:1990} as well as \cite{Thomason:1988}. Children similarly 
go through a single word phase (even though a single 
word for them might be multiple words for us), and 
then slowly start to bootstrap their grammar. 
See  \cite{Tomasello:1991} and \cite{Bates:1991}. 
They are still observed in the very first phases of 
child language acquisition or in pidgins that spontaneously 
form when two communities with widely diverging languages
need to interact. Out of proto-languages, languages 
with a fully-fledged syntax must have emerged at some point. 
How this occurred is still a heavily debated mystery. In this volume 
I only treat the origins of proto-languages.}

Children acquire their first words around the first year
of life.\footnote{Representative work in the study of child language
learning focuses mostly on the acquisition of specific meanings.
See  \cite{Gleitman:1994}, 
\cite{Clark:1993}, \cite{Bowerman:1996}.}
Most people believe that they do this 
as a result of hearing a particular word repeated several times in a
certain context and gradually abstracting an
association between a word and a meaning. But 
how is it that they know which meaning to associate with a 
particular word? How is it that word acquisition goes
so rapid? We will follow a different 
approach, which leaves an open question whether
this applies to human word acquisition as well. The
approach will be selectionist. The agents {\itshape construct} 
hypotheses, either on the basis of one specific case 
where they guess through a non-verbal strategy what 
the meaning of an unknown word might be, 
or they have simply invented a new word because
they do not have one yet. The hypotheses are 
then tried out in subsequent games and either receive confirmation
or are falsified. As a side effect of this local behaviour, 
a global self-organising dynamic process arises leading gradually to coherence.

@@

Through such disambiguating situations, meanings of words
get clarified and the lexicons of the agents
become more similar. Note that the 
dominating meaning of `lu' can still go in 
two directions. For example, {\bfshape a2}
could have used `lu' with {\bfshape a1} 
in a situation where its only possible meaning 
is [\textsc{hpos} 0.0–0.5] (left). Then {\bfshape a1}, now playing the role
of hearer, would have stored the association between
`lu' and [\textsc{hpos} 0.0–0.5]. If this happened often 
enough [\textsc{hpos} 0.0–0.5] (left) might
have become the dominant meaning of `lu', instead of 
[\textsc{vpos} 0.0–0.5] (lower). This shows clearly that the 
evolution of language will never be predictable. At a critical 
bifurcation, small preferential differences between 
the agents or the chance occurrence of certain situations
may tilt the competition in one direction or the other.\footnote{This high sensitivity to initial conditions is 
one of the characteristics of a chaotic dynamic
system, see \cite{Lorenz:1993}.
Indeed, as I will explore in more detail later, 
evolving languages have all the characteristics
of complex adaptive systems, including the 
potential for punctuated equilibria.}
There is no right
or wrong solution in the language game and no one has
any more rights than anyone else.

@@

Another major difference with classical theories of 
meaning and reference is 
that the repertoire of meanings and form-meaning pairs is open-ended
and subject to change at any time. Logicians would say that the 
truth conditions of the forms are non-monotonic. \is{non-monotonicity}
Non-monotonicity is unavoidable in the case of grounded 
situated agents with limited rationality.\footnote{Examples of these processes are described in: 
\cite{Bybee:1994}, \cite{Traugott:1991}.
For attempts to explain these ``grammaticalisation'' 
processes in terms of general cognitive operations, see  
\cite{Heine:1997}.}
Agents should always be allowed to 
introduce new forms or recruit existing forms to express new
meanings, simply because the set of meanings must expand
to cope with novel contexts and new communicative
situations. After enough interactions among the 
members of the same group in a relatively stable 
environment, we expect there to be
a stable set of conventionalised form-meaning 
relations, but it cannot be
expected that everything anyone ever wants to say is
already conventionalised, and so there will always
be turbulence at the fringes. The idea that the lexicon
and the syntax of a language are static
entities is completely false. Both are in constant flux.
Non-monotonicity is consequently one of the big 
topics in logic-based approaches to artificial intelligence.

@@

The spontaneous generation of repertoires of complex 
semantic strategies already explains some characteristics 
that are reflected in full-fledged languages: 
There is hierarchical structure, because one strategy 
may call upon another strategy to achieve a subgoal, 
and a strategy may potentially call upon itself thus introducing 
recursivity. There is also a functional specialisation
of categories. They are now sometimes used for filtering
the members of a set, sometimes for ordering them, sometimes
for modifying another category before it is used for 
filtering, and so on. Hierarchical structuring, recursivity
and functional specialisation therefore do not have to 
originate in language. The fact that we see them in 
language structure is a reflection of the generic 
hierarchical and functional nature 
of semantic strategies.\footnote{The fact that languages change is, of course, 
well-known in linguistics even though it is less a focal 
research topic than used to be in the 19th century. 
See  \cite{McMahon:1994} for a brief introduction into the 
subject. Language change has many of the characteristics
of biological change but it takes place in cultural 
as opposed to biological time. See  \cite{Labov:1994}.}

Second, my hypothesis is that the need for communicating 
more and more complex semantic strategies and the 
conceptual structures they use, has pushed language 
users to start using more and more complex form characteristics, 
starting with word order but also intonation, stress 
patterns, function words, morphological variations of words, 
etc. Here another aspect comes into 
play: the ability to recognise forms and structures, 
which we also need to recognise structured objects in 
scenes for example, and the ability to assemble a 
structure satisfying a set of constraints. The collective
dynamics which are responsible for the propagation and the spontaneous 
self-organisation of coherence in lexicons should also apply  
here. Grammars can be seen as ecologies, where
form-meaning pairs compete in the population. New 
syntactic and semantic categories, new constructions and 
new uses of grammatical conventions are continuously created,
and existing ones may destabilise and become
in disuse.\footnote{The semantic strategies are similar to the 
cognitive processes that have been studied extensively 
in cognitive grammar (\citealt{Langacker:1987}). Formal versions
of some of these processes have been formulated as 
part of Montague grammar (\citealt{Montague:1974}). In 
computational linguistics research, various 
attempts have been made to formalise conceptualisation
and meaning application strategies (\citealt{Gazdar:1989}).}

@@

\enlargethispage{1\baselineskip}
According to this theory, natural language structure is
not the consequence of an 
autonomous innate language device which evolved 
by a genetic mutation or series of mutations. Instead 
cognitive mechanisms and structures already in place have been 
marshalled to get a language system off the ground, even though 
once language became essential for the rapid incorporation
of new individuals and the general organisation of 
activities in human populations, it must have started
to recruit vast amounts of brain capacity and stimulated 
other cognitive faculties, such as categorisation,
episodic memory or problem solving, to become in turn 
much more complex and versatile.\footnote{There has been an ongoing nature versus nurture debate 
with respect to the origins and acquisition of language. 
The different issues in this debate are already well
illustrated in \cite{Piattelli:1980}.}
